Symbolic Weighted Language Models and\\ Quantitative Parsing over Infinite Alphabets

### 2021-10-12 introduction.tex
intro: remarque org.

% !TEX root = main.tex
%
% Introduction
%
Parsing is the problem
\florent{mv}
of structuring a linear representation
(a finite word) according to a language model. % (a formal grammar).
% natural language, programming language,
%
Most of the context-free parsing approaches~\cite{GruneJacobs08parsing}
assume a finite and reasonably small input alphabet. %models and algorithms
Such a restriction makes perfect sense in the context of
NLP tasks such as constituency parsing, 
compilers or interpreters for programming languages.
Considering large or infinite alphabets can however be of
practical interest when dealing with large characters encodings such as UTF-16~\cite{dAntoni21CACM},
% processing strings in \eg for vulnerability detection in Web-applications~\cite{dAntoni21CACM},
%
analysis of data streams, serialization of structured documents~\cite{Segoufin06csl,NevenSchwentickVianu04FSMinfinite},
or processing timed execution traces~\cite{Bouyer03algebraic}.
% algebraic definition of a class of data languages
% (notion of monoid recognizability, based on registers, comparable to Bojancszik et al. data words)
%

The latter case is related to a motivation  of  the present work:
automated music transcription. Representations that capture  music performances
are essentially linear:
audio files or the widely used
MIDI format~\cite{Selfridge-Field97beyondMIDI}. %\cite{MIDIfile}.
Such representations ignore the hierarchical structures that frame the
conception of music, at least in the Western area. These structures, on the other hand,
are present, either explicitly  or implicitly,
in Common Western Music Notation~\cite{Gould11Notation}:
%in music notation~\cite{Gould11Notation}:
Music scores are partitioned in measures,
measures in beats, and beats can be further recursively divided.
It follows that music events do not occur at arbitrary timestamps,
but respect a discrete partitioning of the  timeline incurred by
these recursive divisions.
The \emph{transcription problem} takes
as input a linear representation (audio or MIDI) and aims at re-constructing
these structures
by mapping input events to this hierarchical rhythmic space.
It can therefore be stated as a parsing problem~\cite{foscarin:hal-01988990}
over an infinite alphabet of timed events.

@@

Other extensions of language models  %(automata and grammars)
help in dealing with non-determinism by computing weight values.
With an ambiguous grammar, 
\reviews{1) jump from automata to grammars is uneasy and not really necessary}
there may exist several derivations
(\emph{abstract syntax trees} -- AST) % \ie the result of an analyze.
yielding one input word. % structuring a word,
The association of one weight value %in semiring domain
to each AST permits to select a best one (or $n$ bests). % a fixed number of bests
% = ...a ranking of ASTs
%This is roughly the principle of \emph{weighted parsing} approaches
\florent{introduce pb "weighted parsing" and ref.\cite{MorbitzVogler19weighted-parsing}.} 
In \emph{weighted language models}~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing},
like \eg probabilistic context-free grammars % (CFG),
and weighted automata (\WA) \cite{Droste09handbook},
a weight is associated to each transition rule, % production rule
and the rule's weights can be combined with an
associative product operator~$\otimes$ to yield the weight of an AST.
\reviews{1) AST for grammar, run for WA}
A second operator~$\oplus$
is moreover used to resolve the ambiguity raised by the existence
of several (in general exponentially many) AST
associated to a given input word.
Typically, $\oplus$  selects the best of two weight values.
%Intuitively,~$\oplus$ selects, or ranking, the syntax trees.
The weight domain, equipped with these two operators is typically 
a \emph{semiring} %$\Semiring$
where $\oplus$ can be extended to infinite sums,
%\cite{Eilenberg74automata}
such as the Viterbi semiring and the tropical min-plus algebra% % see Figure~\ref{fig:semirings}.
%of domain $\mathbb{R}^+ \cup \{ +\infty\}$,
%where $\oplus$ is min and $\otimes$ is plus .
%by ranking
%making the weight domain a semiring.
%Some efficient specialized parsing algorithms~\cite{Huang05kbest} have been proposed in this context
%% models represented as hypergraphs \cite{Huang05kbest}
%in order to compute the $n$ best syntax trees of a given input word without having to enumerate them all.
%Generally based on dynamic programming, these algorithms rely on
%additional algebraic properties of~$\Semiring$.
%-- see \eg~\cite{Huang05kbest} for some NLP applications.
%The extraction of $n$ best list is useful
%the problem: quantitative parsing or symbolic parsing
%of parsing of words over infinite input alphabet.

\florent{Both WA and SA generalized \cite{Herrmann16dlt,Herrmann20phd}...
         we present results...}
In this paper, we present 
a framework for weighted parsing over infinite input alphabets.
It is based on \emph{symbolic weighted} finite states language models (\swM),
generalizing the Boolean guards of~\SA %in transitions
to functions into an arbitrary semiring,
and generalizing also~\WA, by handling infinite alphabets (Figure~\ref{fig:hierarchy}).
\reviews{cite \cite{Herrmann16dlt,Herrmann20phd}}
%
In short, a transition rule $q \xrightarrow{\phi} q'$, from state $q$ to $q'$,
%\florent{c'est un résumé de la Fig. \ref{fig:hierarchy}}
%input symbols are variables, %(parameters)
is labeled by a function~$\phi$ associating to every input symbol $a$ a weight value $\phi(a)$
in a semiring $\Semiring$.

@@

\florent{Moreover, application parsing. believe results appliable to other decision pb.}

>>>>

% !TEX root = main.tex
%
% Introduction
%
Parsing is the problem
\florent{deplacer plus loin (application)}
of structuring a linear representation
(a finite word) according to a language model. % (a formal grammar).
% natural language, programming language,
%
Most of the context-free parsing approaches~\cite{GruneJacobs08parsing}
assume a finite and reasonably small input alphabet. %models and algorithms
Such a restriction makes perfect sense in the context of
NLP tasks such as constituency parsing, 
compilers or interpreters for programming languages.
Considering large or infinite alphabets can however be of
practical interest when dealing with large characters encodings such as UTF-16~\cite{dAntoni21CACM},
% processing strings in \eg for vulnerability detection in Web-applications~\cite{dAntoni21CACM},
%
analysis of data streams, serialization of structured documents~\cite{Segoufin06csl,NevenSchwentickVianu04FSMinfinite},
or processing timed execution traces~\cite{Bouyer03algebraic}.
% algebraic definition of a class of data languages
% (notion of monoid recognizability, based on registers, comparable to Bojancszik et al. data words)
%

The latter case is related to a motivation  of  the present work: \florent{idem}
automated music transcription. Representations that capture  music performances
are essentially linear:
audio files or the widely used
MIDI format~\cite{Selfridge-Field97beyondMIDI}. %\cite{MIDIfile}.
Such representations ignore the hierarchical structures that frame the
conception of music, at least in the Western area. These structures, on the other hand,
are present, either explicitly  or implicitly,
in Common Western Music Notation~\cite{Gould11Notation}:
%in music notation~\cite{Gould11Notation}:
Music scores are partitioned in measures,
measures in beats, and beats can be further recursively divided.
It follows that music events do not occur at arbitrary timestamps,
but respect a discrete partitioning of the  timeline incurred by
these recursive divisions.
The \emph{transcription problem} takes
as input a linear representation (audio or MIDI) and aims at re-constructing
these structures
by mapping input events to this hierarchical rhythmic space.
It can therefore be stated as a parsing problem~\cite{foscarin:hal-01988990}
over an infinite alphabet of timed events.

@@

Other extensions of language models  %(automata and grammars)
help in dealing with non-determinism by computing weight values.
\florent{quantitative reasoning about programs}
With an ambiguous grammar, 
\reviews{1) jump from automata to grammars is uneasy and not really necessary}
there may exist several derivations
(\emph{abstract syntax trees} -- AST) % \ie the result of an analyze.
yielding one input word. % structuring a word,
The association of one weight value %in semiring domain
to each AST permits to select a best one (or $n$ bests). % a fixed number of bests
% = ...a ranking of ASTs
%This is roughly the principle of \emph{weighted parsing} approaches
\florent{introduce pb "weighted parsing" and ref.\cite{MorbitzVogler19weighted-parsing}.} 
In \emph{weighted language models}~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing},
like \eg probabilistic context-free grammars % (CFG),
and weighted automata (\WA) \cite{Droste09handbook},
a weight is associated to each transition rule, % production rule
and the rule's weights can be combined with an
associative product operator~$\otimes$ to yield the weight of an AST.
\reviews{1) AST for grammar, run for WA}
A second operator~$\oplus$
is moreover used to resolve the ambiguity raised by the existence
of several (in general exponentially many) AST
associated to a given input word.
Typically, $\oplus$  selects the best of two weight values.
%Intuitively,~$\oplus$ selects, or ranking, the syntax trees.
The weight domain, equipped with these two operators is typically 
a \emph{semiring} %$\Semiring$
where $\oplus$ can be extended to infinite sums,
%\cite{Eilenberg74automata}
such as the Viterbi semiring and the tropical min-plus algebra% % see Figure~\ref{fig:semirings}.
%of domain $\mathbb{R}^+ \cup \{ +\infty\}$,
%where $\oplus$ is min and $\otimes$ is plus .
%by ranking
%making the weight domain a semiring.
%Some efficient specialized parsing algorithms~\cite{Huang05kbest} have been proposed in this context
%% models represented as hypergraphs \cite{Huang05kbest}
%in order to compute the $n$ best syntax trees of a given input word without having to enumerate them all.
%Generally based on dynamic programming, these algorithms rely on
%additional algebraic properties of~$\Semiring$.
%-- see \eg~\cite{Huang05kbest} for some NLP applications.
%The extraction of $n$ best list is useful
%the problem: quantitative parsing or symbolic parsing
%of parsing of words over infinite input alphabet.

\florent{Both WA and SA generalized by \cite{Herrmann16dlt,Herrmann20phd}...
         we present results...}
In this paper, we present 
a framework for weighted parsing over infinite input alphabets.
It is based on \emph{symbolic weighted} finite states language models (\swM),
generalizing the Boolean guards of~\SA %in transitions
to functions into an arbitrary semiring,
and generalizing also~\WA, by handling infinite alphabets (Figure~\ref{fig:hierarchy}).
\reviews{cite \cite{Herrmann16dlt,Herrmann20phd}}
%
In short, a transition rule $q \xrightarrow{\phi} q'$, from state $q$ to $q'$,
%\florent{c'est un résumé de la Fig. \ref{fig:hierarchy}}
%input symbols are variables, %(parameters)
is labeled by a function~$\phi$ associating to every input symbol $a$ a weight value $\phi(a)$
in a semiring $\Semiring$.

@@

\florent{Application to parsing and tanscription. these results are appliable to other decision pb.}
